Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-12 surfaces-write-up/surface-write-up.tex
spacing & name change hirzebruch section

\section{Canonical Rings on \texorpdfstring{$F_k$}{}}
Let $D=\sum_{i=1}^m \alpha_i C_i$ for some curve $C_i\in F_k$, such that $C_1, \ldots, C_4$ are independent curves with bi-degrees (1,0), (1,0), (0,1), and (0,1) respectively.  Let $C_i = V(f_i)$ where $f_i \in \mathscr{O}(a_i, b_i)$.  Then each $C_i$ corresponds to a polynomial $f_i$ where $f_1$ and $f_2$ are independent linear polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent linear polynomials in $y_0$ and $y_1$.
\begin{defn}
Suppose $D = \sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(F_k)$.  Then define 
\begin{equation}\label{eqn:define-T=(D)}
	T_=(D) = \left\{i \in \{1, \ldots, m\}: a_i \sum_{k=1}^m b_k \alpha_k = b_i \sum_{k=1}^m a_k \alpha_k \right\},
\end{equation}
\begin{equation}\label{eqn:define-T+(D)}
	T_+(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} a_i \sum_{k=1}^m \alpha_k b_k > b_i \sum_{k=1}^m \alpha_k a_k \right\},
\end{equation}
and
\begin{equation}\label{eqn:define-T-(D)}
	T_-(D) = \left\{ j \in \{1, \ldots, m\}: a_j \sum_{k=1}^m \alpha_k b_k < b_j \sum_{k=1}^m \alpha_k a_k s \right\}.
\end{equation}
\end{defn}

\begin{lem}
$R_D$ is generated in degree at most
\[
	\sum_{i\in T_=(D)} \gcd(a_i, b_i)\ell_i + \sum_{i\in T_+(D) \atop j\in T_-(D)} (a_i b_j- a_j b_i)\ell_{i,j}.
\]
\end{lem}
\begin{proof}
Let $(a_i, b_i)$ be the bi-degrees of $f_i$.

Notice that if $\gamma \in (R_D)_d$, then 
\[
	g = \sum_{i=1}^m {f_i}^{c_i}
\]
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^m c_i a_i = 0$, and $\sum_{i=1}^m c_i b_i = 0$.  Then we can view $g$ inside of the lattice
$S\subseteq \mathbb{R}^{n+1}$ given by ordered elements of the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$ for all $i$, $\sum_{i=1}^m c_i a_i = 0$, and $\sum_{i=1}^m c_i b_i = 0$. 
From here, determining a spanning set for $(R_D)_d$ reduces to finding the extremal rays of $S$ \todo{cite previous lemma}.

To do this, we extend the method of O'Dorney \todo{cite}.  We first view the cone $T\subseteq \mathbb{R}^{n+1}$ given by elements of the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -\alpha_i d$ for all $i$ and $\sum_{i=1}^n c_i (a_i + b_i) =0$.  By O'Dorney, this has extremal rays given by 
\[
	e_i = (1, -\alpha_1, \ldots, -\alpha_{i-1}, \frac{\sum_{j \ne i} \alpha_j(a_j+ b_j)}{a_i + b_i}, -\alpha_{i+1}, \ldots, - \alpha_m).
\]
Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_m$.  We can then intersect the hyperplane $H$ given $\sum_{i=1}^m  a_i X_i = 0$.  The extremal rays of $S$ are given by the set of $e_i$ contained in $H$ together with the points of intersection $e_{i,j}$ of $H$ and the line segment connecting $e_i, e_j \not \in H$.

@@

For $i \in T_=(D)$
define
\[
	\epsilon_i := (\prod_{k \ne i} f_k^{-\alpha \ell_i \gcd(a_i, b_i)}) (f_i^{\ell_i \frac{\gcd(a_i, b_i)}{a_i + b_i}\sum_{k \ne i} \alpha_k (a_k + b_k)}) \in H^0(\ell_i \gcd(a_i, b_i) D).
\]
For $i \in T_+(D)$ and $j \in T_-(D)$, define
\[
	\epsilon_{i, j} = (\prod_{k = 1}^m f_k^{-\alpha \ell_{i,j} (a_i b_j - a_j b_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i,j}(a_i b_j - a_j b_i)D)
\]
where
\[
	s_1 = b_j \sum_{k \ne i,j} \alpha_k a_k - a_j \sum_{k\ne i, j} \alpha_k b_k
\]
and
\[
	s_2 = b_i \sum_{k \ne i,j} \alpha_k a_k - a_i \sum_{k \ne i, j} \alpha_k b_k.
\]
Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \todo{cite earlier proposition}, so by applying that proposition $R_D$ is generated in degree at most
\[
	\sum_{i\in T_=(D)} \gcd(a_i, b_i)\ell_i + \sum_{i\in T_+(D) \atop j\in T_-(D)} (a_i b_j- a_j b_i)\ell_{i,j}.
\]
\end{proof}

\todo{factor $\phi$ through $\psi$ and $\chi$.}
\begin{lem}
$\ker(\psi)$ is generated in degree at most \todo{insert number}.
\end{lem}
\begin{proof}
We first claim that $\psi$ has a relation of the form
\[
	(y_j - \beta_j)\prod_{i=1}^m {y_i}^{c_{i}}
\]
whenever $d\in \mathbb{N}$, $c_i \ge -\alpha_i d$, and $\beta_j \in k[y_1, \ldots, y_4]$ satisfies $\psi(\beta_j)\in \mathscr{O}(a_j, b_j)$.
To find such a $\beta$, we simply write $\epsilon_j$ as a polynomial in $\epsilon_1, \ldots \epsilon _4$.

@@

\section{Acknowledgments}

We are grateful to David Zureick-Brown for introducing us to this
subject, for providing invaluable guidance,
and for his mentorship. We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.

>>>>

\section{Canonical Rings on Hirzebruch surfaces}
Let $D=\sum_{i=1}^m \alpha_i C_i$ for some curve $C_i\in F_k$, such
that $C_1, \ldots, C_4$ are independent curves with bi-degrees $(1,0)
, (1,0), (0,1)$, and $(0,1)$ respectively. Let $C_i = V(f_i)$ where
$f_i \in \mathscr{O}(a_i, b_i)$. Then each $C_i$ corresponds to a
polynomial $f_i$ where $f_1$ and $f_2$ are independent linear
polynomials in $x_0$ and $x_1$ and $f_3$ and $f_4$ are independent
linear polynomials in $y_0$ and $y_1$.

\begin{defn}
Suppose $D = \sum_{i=1}^m \alpha_i C_i \in \mathbb{Q} \otimes \di(F_
k)$. Then define 
\begin{equation}\label{eqn:define-T=(D)}
	T_=(D) = \left\{i \in \{1, \ldots, m\}: a_i \sum_{k=1}^m b_k 
\alpha_k = b_i \sum_{k=1}^m a_k \alpha_k \right\},
\end{equation}

\begin{equation}\label{eqn:define-T+(D)}
	T_+(D) = \left\{ i \in \{1, \ldots, m\}:  \vphantom{\sum_{k=1}^m} 
a_i \sum_{k=1}^m \alpha_k b_k > b_i \sum_{k=1}^m \alpha_k a_k 
\right\},
\end{equation}

\noindent
and
\begin{equation}\label{eqn:define-T-(D)}
	T_-(D) = \left\{ j \in \{1, \ldots, m\}: a_j \sum_{k=1}^m \alpha_
k b_k < b_j \sum_{k=1}^m \alpha_k a_k s \right\}.
\end{equation}
\end{defn}

\begin{lem}
$R_D$ is generated in degree at most
\[
	\sum_{i\in T_=(D)} \gcd(a_i, b_i)\ell_i + \sum_{i\in T_+(D) \atop j\in T_-(D)} (a_i b_j- a_j b_i)\ell_{i,j}.
\]
\end{lem}

\begin{proof}
Let $(a_i, b_i)$ be the bi-degrees of $f_i$. Notice that if $\gamma
\in (R_D)_d$, then 
\[
	g = \sum_{i=1}^m {f_i}^{c_i}
\]

\noindent
where each $c_i \ge - \alpha_i c_i$, $\sum_{i=1}^m c_i a_i = 0$,
and $\sum_{i=1}^m c_i b_i = 0$. Then we can view $g$ inside of the
lattice $S \subseteq \mathbb{R}^{n + 1}$ given by ordered elements of
the form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge - d \alpha_i$
for all $i$, $\sum_{i=1}^m c_i a_i = 0$, and $\sum_{i = 1}^m c_i b_i =
 0$. From here, determining a spanning set for $(R_D)_d$ reduces to
finding the extremal rays of $S$ \todo{cite previous lemma}.

To do this, we extend the method of O'Dorney \todo{cite}. We first
view the cone inside of $\mathbb{R}^{n+1}$ given by elements of the
form $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -\alpha_i d$ for
all $i$ and $\sum_{i = 1}^n c_i (a_i + b_i) =0$. By O'Dorney, this
has extremal rays given by

\[
	e_i = (1, -\alpha_1, \ldots, -\alpha_{i-1}, \frac{\sum_{j \ne i}
	\alpha_j(a_j+ b_j)}{a_i + b_i}, -\alpha_{i + 1}, \ldots, - \alpha_m).
\]

Let $\Delta^{n-1}$ be the simplex with vertices $e_1, \ldots e_m$.
We can then intersect the hyperplane $H$ given $\sum_{i=1}^m a_i X_i
= 0$ to get a subspace $S$. The extremal rays of $S$ are given
by the set of $e_i$ contained in $H$ together with the points of
intersection $e_{i,j}$ of $H$ and the line segment connecting $e_i,
e_j \not \in H$.

@@

For $i \in T_=(D)$
define
\[
	\epsilon_i := (\prod_{k \ne i} f_k^{-\alpha \ell_i \gcd(a_i, b_i)})
	(f_i^{\ell_i \frac{\gcd(a_i, b_i)}{a_i + b_i}\sum_{k \ne i}
	\alpha_k (a_k + b_k)}) \in H^0(\ell_i \gcd(a_i, b_i) D).
\]

For $i \in T_+(D)$ and $j \in T_-(D)$, define
\[
	\epsilon_{i, j} := (\prod_{k = 1}^m f_k^{-\alpha \ell_{i,j} (a_i b_
j - a_j b_i)}) {f_i}^{s_1} {f_j}^{s_2} \in H^0(\ell_{i,j}(a_i b_j - 
a_j b_i)D)
\]

where
\[
	s_1 = b_j \sum_{k \ne i,j} \alpha_k a_k - a_j \sum_{k\ne i, j}
	\alpha_k b_k
\]

\noindent
and
\[
	s_2 = b_i \sum_{k \ne i,j} \alpha_k a_k - a_i \sum_{k \ne i, j}
	\alpha_k b_k.
\]

Let $E$ be the set of $\epsilon_i$ and $\epsilon_{i,j}$ given 
above, satisfying the respective conditions.
$E$ satisfies the conditions of proposition \todo{cite earlier 
proposition}, so by applying that proposition $R_D$ is generated in 
degree at most
\[
	\sum_{i\in T_=(D)} \gcd(a_i, b_i)\ell_i + \sum_{i\in T_+(D) \atop
	j \in T_-(D)} (a_i b_j- a_j b_i)\ell_{i,j}.
\]
\end{proof}

\todo{factor $\phi$ through $\psi$ and $\chi$.}
\begin{lem}
$\ker(\psi)$ is generated in degree at most \todo{insert number}.
\end{lem}

\begin{proof}
We first claim that $\psi$ has a relation of the form
\begin{equation}\label{eqn:relations-psi}
	(y_j - \beta_j)\prod_{i=1}^m {y_i}^{c_{i}}
\end{equation}

\noindent
for some $d\in \mathbb{N}$, $c_i \ge -\alpha_i d$, $a_j + \sum_{i = 1}
^m a_i c_i = 0$, $b_j + \sum_{i=1}^m b_i c_i = 0$, and $\beta_j \in
k[y_1, \ldots, y_4]$ satisfies $\psi(\beta_j)\in \mathscr{O}(a_j,
b_j)$.

To find such a $\beta$, we simply write $\epsilon_j$ as a 
polynomial in $\epsilon_1, \ldots \epsilon _4$. Furthermore, these
generate all relations, since they allow us to reduce any $\prod_{i =
1}^n \epsilon_i^{r_i}$ to a canonical form. \todo{add reference to
earlier description of canonical form.}

Note that we can view the relations from Equation
\ref{eqn:relations-psi} as an element of $\mathbb{R}^{n+1}$ given
by $(d, c_1, \ldots, c_n)$.  Let $S'\subseteq \mathbb{R}^{+}$ be the
set of $(d, c_1, \ldots, c_n)$ satisfying $c_i \ge -\alpha_i d$,
$a_j + \sum_{i = 1}^m a_i c_i = 0$, and $b_j + \sum_{i=1}^m b_i c_i
= 0$.  Note that $S'$ is a translation of $S$ by \todo{finish}

@@

\section{Acknowledgments}
We are grateful to David Zureick-Brown for introducing us to this
subject, for providing invaluable guidance,
and for his mentorship. We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.
